The Small World Network Effect in Software Project Teams \author{Kevin Peterson\\ \small \texttt{pete1968@umn.edu} }

### 2013-12-01 paper/main.tex
Rearranged and shuffled.

\begin{abstract}
Team cohesion and the dynamics of team forming are important parts of any project, with software projects being no exception. An interesting aspect of team building is the relationships formed between the team members. If follows that visualizing software team members as a graph can give some insight as to what the optimal team conditions are. As team members move between projects, these graphs become more and more connected as team members collaborate and form relationships. We show that this connectivity, known as the ``small world effect," has a positive impact on team performance when the connectivity levels are moderate. Performance degrades, however, at both very high and very low levels of connectivity. This aligns with similar research findings of non-software teams.
\end{abstract}

@@

\section{Introduction}
A social network is a graph of people and the connections between them. The dynamics of social networks have proved to be an important research topic for many different areas of study, from academic publication\cite{barabasi2002evolution}, to the success of broadway musicals\cite{uzzi2005collaboration}. In general, a social network is important to understanding how ideas and influence are spread\cite{kempe2003maximizing}, from which we can investigate the best conditions for optimal group perfomance within these networks.

A `small world' network is a social network characterized by high clustering of graph nodes paired with a short average path between nodes\cite{watts1998collective}. `Clustering' in this context means how closely nodes in the graph are related to each other, or "the friend of my friend is also my friend." Milgram, in is seminal study, found that any two people are linked through a chain of friends and acquaintances on average of 6 people long\cite{milgram1967small}, or in other words, the average `path length' between any two people.

@@

We follow Uzzi's work, and expand it to software teams. We hypothesize that the qualities of a small world graph that foster collaboration, creativity, and the efficiency of how knowledge is transfered\cite{latora2001efficient} will also apply to software project teams. In this work, we investigate the `small world' phenomenon quantitatively by exploring open source project data, and qualitatively by a series of interviews of subject matter experts, and look to compare Uzzi's findings to ours.

\subsection{Hypothesis}
The formalization of the `small world' quality of a graph can be expressed by the \textit{Small World Quotient}, or $Q$\cite{watts1999small,watts1998collective}. The value $Q$ is calculated by dividing the clustering factor of graph by the average path length between nodes. Uzzi explored this value in graphs of Broadway productions\cite{uzzi2005collaboration}. His findings suggest that the success of Broadway productions is greatest when teams have a good mix of new and familiar members. Productions are less likely to be successful when team members are very familiar with each other. Also, probability of success drops much the same way when there is little familiarity between team members. This indicates that there is an optimal ratio of new and familiar members on a team -- or, an optimal value of $Q$ when related to performance.

@@

\subsection{Analysis}
The collected data was organized into a graph structure, with the nodes representing the collaborators and the edges representing one or more shared project between collaborators. Various properties of the graph are analyzed, followed by a qualitative section to help link the quantitative analysis to subject matter expert experiences.

@@

The Freecode data can be represented as a bipartite graph. A bipartite graph is a specialized type of graph containing two disjoint sets of verticies. This can be denoted by ${G=(P,C,E)}$, where ${P \cap C = \emptyset}$. For our purposes, let set $P$ be the set of all projects, and set $C$ be the set of all contributors. Each project node, therefore, is only connected to contributor nodes, and vice versa, as show in figure \ref{fig:example_bipartite_graph}. This bipartite representation is typical of what is found in social and collaboration networks\cite{ramasco2004self}.

@@

We assume for throughout that any graph exhibiting {$Q > 1$} is a small-world graph.

@@

For both datasets, we base our performance metric on \textit{use}. For Freecode, we use the \textit{Popularity Score} metric, which is calculated the following formula\footnote{http://help.freecode.com/kb/statistics/how-do-you-measure-a-projects-popularity}:

@@

For the SourceForge dataset, performance is based on the repository's internal \textit{rank} metric. Note that there was no attempt to normalize the two performance metrics between datasets. Because of this, no cross-dataset measurements can be made, as the performance metric is only valid in the context of the enclosing dataset.

@@

Summary statistics for the graph analysis are captured in figure \ref{fig:summary_stats}.  Statistics analyzed included:\\
$Q$ - \textit{Small World Quotient}\\
$C^\Delta$ - \textit{Clustering Coefficient - Transitivity}\\
$\overline{C^\lambda}$ - \textit{Clustering Coefficient - Average}\\
$L$ - \textit{Average Path Length}\\
$N$ - \textit{Number of Nodes}\\
$E$ - \textit{Number of Edges}\\
As measured over the number of projects $P$, with the total number of contributors $C$. The number of connected subgraphs analyzed is shown as $P_c$.

A mean value of $Q$ was observed to be $> 1$ for both data sets, which meets the criteria for a small-world network\cite{humphries2008network}. The median, however, is 1.0 in both cases, which does not match the small world network criteria. A possible explanation of this is the power law distribution of $Q$ observed in both data sets (figures \ref{fig:q_fc_distribution} and \ref{fig:q_sf_distribution}. This indicates that a large proportion of $Q$ values are around 1.\\

$C^\Delta$ and $\overline{C^\lambda}$ both indicate values approaching 1. This observation indicates a high degree of clustering\cite{watts1998collective}. hat any indicates that the majority of subgraphs have a $Q$ and $C^\Delta$ value of close to either 0 or 1. Exploring this phenomenon further, we can derive that most graphs resemble the middle or the bottom graph in figure \ref{fig:example_graphs}, with relatively few resembling the top graph.\\

Note that $P_c$ values are relatively small compared to $P$ and $C$ values. It would be intuitive to assume that {$C \times P_c \approx P$} -- in other words, if we split the entire graph into connected subgraphs, the number of subgraphs multiplied by the average number of nodes (contributors) per subgraph should equal the total number of contributors ($C$). This, however, is not the case. In fact, $P_c$ is notably smaller than expected. Recall that during analysis, subgraphs of {$< 3$} nodes were not considered. We can assume then that a large number of subgraphs are very small, with only one or two nodes. Further analysis is needed to fully understand the implications of this.

@@

Small-world-ness, or $Q$, is generally between 1 and 3 for most sub-graphs.

\subsection{Qualitative}
The research interviews provided a look into how some of these quantitative findings translate into actual industry practice. It also helped us to get a deeper understanding of what we had found in the data analysis, as we could begin to build some workplace context into our results. There were are few re-occurring themes from which we will base our analysis\cite{stmartin_interview,rana_interview}:

@@

\noindent\\\textit{\textbf{Teams can stagnate without new members.}}\\
Both interview participants indicated that new members were essential in bringing new ideas to the team. A highly static team (or, a team with a high $Q$ value), runs the risk of falling behind the technology trends.

@@

\section{Future Research}
Social networks, interpersonal relationships, and software development are all complex topics. We have explored a narrow slice of these in context, and there are some important areas of future exploration. First, the contributor graph contains a large amount of subgraphs, and as a whole appears less clustered than some other observed social network graphs -- consistent with Madey's findings\cite{madey2002open} on SourceForge data. As noted in our analysis, a large number of these subgraphs have only one or two contributors. This implies that a large number of contributors work alone and on a single project. We do not, with the current analysis, know how (or if) this degrades the quality of our findings.\\

>>>>

\begin{abstract}
Team cohesion and the dynamics of team forming are important parts of any project, with software projects being no exception. An interesting aspect of team building is the relationships formed between the team members. It follows then that visualizing software team members as a graph may be a natural way to explore these relationships. As team members move between projects, these graphs become more and more connected as team members collaborate and form relationships. We show that this connectivity, known as the ``small world effect," has a positive impact on team performance when the connectivity levels are moderate. Performance degrades, however, at both very high and very low levels of connectivity. This aligns with similar research findings of non-software teams.
\end{abstract}

@@

\section{Introduction}
A social network is a graph of people and the connections between them. The dynamics of social networks have proved to be an important research topic for many different areas of study, from academic publication\cite{barabasi2002evolution}, to the success of Broadway musicals\cite{uzzi2005collaboration}. In general, a social network is important to understanding how ideas and influence are spread\cite{kempe2003maximizing}. Given that, we can begin to investigate the best conditions for optimal group performance within these networks.

A ``small world'' network is a social network characterized by high clustering of graph nodes paired with a short average path between nodes\cite{watts1998collective}. ``Clustering'' in this context means how closely nodes in the graph are related to each other, or ``the friend of my friend is also my friend.'' Milgram, in is seminal study, found that any two people are linked through a chain of friends and acquaintances on average of 6 people long\cite{milgram1967small}, or in other words, the average ``path length'' between any two people.

@@

We follow Uzzi's work, and expand it to software teams. We hypothesize that the qualities of a small world graph that foster collaboration, creativity, and the efficiency of how knowledge is transfered\cite{latora2001efficient} will also apply to software project teams. In this work, we investigate the ``small world'' phenomenon quantitatively by exploring open source project data, and qualitatively by a series of interviews of subject matter experts, and look to compare Uzzi's findings to ours.

\subsection{Hypothesis}
The formalization of the ``small world'' quality of a graph can be expressed by the \textit{Small World Quotient}, or $Q$\cite{watts1999small,watts1998collective}. The value $Q$ is calculated by dividing the clustering factor of graph by the average path length between nodes. Uzzi explored this value in graphs of Broadway productions\cite{uzzi2005collaboration}. His findings suggest that the success of Broadway productions is greatest when teams have a good mix of new and familiar members. Productions are less likely to be successful when team members are very familiar with each other. Also, probability of success drops much the same way when there is little familiarity between team members. This indicates that there is an optimal ratio of new and familiar members on a team -- or, an optimal value of $Q$ when related to performance.

@@

\subsection{Analysis}
The collected data was organized into a graph structure, with the nodes representing the contributors and the edges representing one or more shared project between contributors. For our purposes, ``contributors'' includes any person that has been identified as being associated with the project by the project hosting site (Freecode and SourceForge, for our purposes). It is important to note that contributor does not always imply code contributions -- as issue reporting and documentation, among other things, are valid contributions. Various properties of the graph were analyzed, followed by a qualitative section to help link the quantitative analysis to subject matter expert experiences.

@@

The contributor/project data can be represented as a bipartite graph. A bipartite graph is a specialized type of graph containing two disjoint sets of verticies. This can be denoted by ${G=(P,C,E)}$, where ${P \cap C = \emptyset}$. For our purposes, let set $P$ be the set of all projects, and set $C$ be the set of all contributors. Each project node, therefore, is only connected to contributor nodes, and vice versa, as show in figure \ref{fig:example_bipartite_graph}. This bipartite representation is typical of what is found in social and collaboration networks\cite{ramasco2004self}.

@@

We assume for throughout that any graph exhibiting {$Q > 1$} is a small world graph.

@@

For both datasets, we base our performance metric on \textit{use}. For Freecode, we use the \textit{Popularity Score} metric, which is calculated by the following formula\footnote{http://help.freecode.com/kb/statistics/how-do-you-measure-a-projects-popularity}:

@@

For the SourceForge dataset, performance is based on the repository's internal \textit{rank} metric. This metric ranks each project from 1 to the total number of projects in SourceForge ($P_{sf}$), or ${1 <= rank <= |P_{sf}|}$. At the time of this writing, no formula for the \textit{rank} metric could be located for citation.  Note that there was no attempt to normalize the two performance metrics between datasets. Because of this, no cross-dataset measurements can be made, as the performance metric is only valid in the context of the enclosing dataset.

@@

Summary statistics for the graph analysis are captured in figure \ref{fig:summary_stats}.  Statistics analyzed included:\\
$Q$ - \textit{Small World Quotient}\\
$C^\Delta$ - \textit{Clustering Coefficient - Transitivity}\\
$\overline{C^\lambda}$ - \textit{Clustering Coefficient - Average}\\
$L$ - \textit{Average Path Length}\\
$N$ - \textit{Number of Nodes}\\
$E$ - \textit{Number of Edges}\\
As measured over the number of projects $P$, with the total number of contributors $C$. The number of connected subgraphs analyzed is shown as $P_c$. The connected subgraphs are the main unit of analysis, and each statistical operation is applied to the set of subgraphs $P_c$.

A mean value of $Q$ was observed to be $> 1$ for both data sets, which meets the criteria for a small-world network\cite{humphries2008network}. The median, however, is 1.0 in both cases, which does not match the small world network criteria. A possible explanation of this is the power law distribution of $Q$ observed in both data sets (figures \ref{fig:q_fc_distribution} and \ref{fig:q_sf_distribution}). This indicates that a large proportion of $Q$ values are around 1.\\

$C^\Delta$ and $\overline{C^\lambda}$ both indicate values approaching 1. This observation indicates a high degree of clustering\cite{watts1998collective}. Exploring this phenomenon further, we can derive that most graphs resemble the middle or the bottom graph in figure \ref{fig:example_graphs}, with relatively few resembling the top graph.\\

Note that $P_c$ values are relatively small compared to $P$ and $C$ values. It would be intuitive to assume that {$N \times P_c \approx C$} -- in other words, if we split the entire graph into connected subgraphs, the number of subgraphs ($P_c$) multiplied by the average number of nodes ($N$) per subgraph should equal the total number of contributors ($C$). This, however, is not the case. In fact, $P_c$ is notably smaller than expected. Recall that during analysis, subgraphs of {$< 3$} nodes were not considered. We can assume then that a large number of subgraphs are very small, with only one or two nodes. Further analysis is needed to fully understand the implications of this.

@@

\subsection{Qualitative}
The research interviews provided a look into how some of these quantitative findings translate into actual industry practice. It also helped us to obtain a deeper understanding of what we had found in the data analysis, as we could begin to build some workplace context into our results. There were are few re-occurring themes from which we will base our analysis\cite{stmartin_interview,rana_interview}:

@@

\noindent\\\textit{\textbf{Teams can stagnate without new members.}}\\
Both interview participants indicated that new members were essential in bringing new ideas to the team. A highly static team (or, a team with a high $Q$ value), runs the risk of falling behind the technology trends. Occasional new members (or, intermediate values of $Q$) was perceived as a positive in terms of information flow

@@

\section{Discussion and Future Research}
Social networks, interpersonal relationships, and software development are all complex topics. We have explored a narrow slice of these in context, and there are some important areas of future exploration. First, the contributor graph contains a large amount of subgraphs, and as a whole appears less clustered than some other observed social network graphs -- consistent with Madey's findings\cite{madey2002open} on SourceForge data. As noted in our analysis, a large number of these subgraphs have only one or two contributors. This implies that a large number of contributors work alone and on a single project. We do not, with the current analysis, know how (or if) this degrades the quality of our findings.\\
